---
title: "Homework 3"
author: "Claire Timm"
date: "2026-03-01"
format: html
editor: visual
---

# Part 1. Set up tasks

```{r}
#| label: packages and data
#| message: false
#| warning: false

#read in packages here
library(tidyverse)
library(here)
library(janitor)
library(readxl)
library(ggeffects)

#read in data here
salinity <- read_csv("data/salinity-pickleweed.csv")
```

# Part 2. Problems

## Problem 1. Slough soil salinity

### a.

To determine the strength of the relationship between salinity (electrical conductivity in mS/cm) and California pickleweed biomass (g), we can run a Pearson’s correlation and a Spearman’s rank correlation. A Pearson’s correlation test allows us to see how two continuous variables are related to each other and what strength (strong, moderate, weak, or none) this relationship is. A Spearman rank correlation measures the strength of a monotonic relationship between variables based on ranked values, this test doesn’t require normality so it stays accurate even with outliers.

### b.

```{r}
#| label: salinity and biomass plot
#| message: false
#| warning: false

#ggplot base 
ggplot(data = salinity, #starting data frame
       aes(x = salinity_mS_cm, #makes x-axis salinity
           y = pickleweed)) + #makes y-axis biomass
  geom_point(color = "deeppink4") + #adds data points in green
  labs(x = "Soil salinity (mS/cm)", #adds label to x-axis
       y = "Pickleweed biomass (g)", #adds label to y-axis
       title = "As soil salinity increases, so does pickleweed biomass") + #adds graph title 
  theme_minimal() #changes graph theme
```

### c.

#### Check assumptions:

```{r}
#| label: checking assumptions
#| message: false
#| warning: false

salinity_model <- lm( #create object to store information
  pickleweed ~ salinity_mS_cm, #formula: response ~ predictor variable
  data = salinity) #data frame used

summary(salinity_model) #display summary

#base R residuals
par(mfrow = c(2, 2)) #create a 2x2 output gride for diagnostic plots
plot(salinity_model) #plot salinity_model information
```

The assumptions I checked for a Pearson correlation model began with a Q-Q Residuals plot assesing normally distributed residuals using QQ plots, here it shows the quantiles following a straight line of the theoretical normal quantiles with only a slight deviation at the end, proving to be normally distributed of residuals. The Residuals vs Fitted plot this checks for homoscedasticity (constant variance) which is proven due to the residuals being somewhat random and evenly spread around the horizontal dotted line, the Scale-Location plot assesses the same but uses the square root of the standardized residuals and this plot also proved there to be constant variance. Lastly in Residuals vs Leverage, this plot determines if there are any outliers influencing the model with Cook's distance, since there aren't any points outside the dotted lines this means there aren't any outliers that would influence the model, which meets all the assumptions for Pearson's correlation.

#### Run test:

```{r}
#| label: run correlation tests
#| message: false
#| warning: false

cor.test(salinity$pickleweed, salinity$salinity_mS_cm,
         method = "pearson")
```

### d.

To determine the strength of the relationship between pickleweed biomass and soil salinity, I used a Pearson’s correlation test since these two continuous variables are related to each other and I wanted to evluate the strength of this relationship and all the assumptions were met for this test. I found a moderate positive relationship between soil salinity and pickleweed biomass (Pearson's r = 0.53, t(21) = 2.9, p = 0.01, $\alpha$ = 0.05).

### e.

### f.

## Problem 2. Personal data

### a.

### b.

## Problem 3. Affective visualization

### a.

### b.

### d.

### e.

## Problem 4. Statistical critique

#### a.

### b.

### c.

### d.
