---
title: "homework-03-script"
author: "Claire Timm"
date: "2026-03-01"
format: html
editor: visual
message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

# Part 1. Set up tasks

```{r}
#| label: packages and data

#read in packages here
library(tidyverse)
library(here)
library(janitor)
library(readxl)

#read in data here
salinity <- read_csv("data/salinity-pickleweed.csv")
```

# Part 2. Problems

## Problem 1. Slough soil salinity

### a.

To determine the strength of the relationship between salinity (electrical conductivity in mS/cm) and California pickleweed biomass (g), we can run a Pearson’s correlation and a Spearman’s rank correlation. A Pearson’s correlation test allows us to see how two variables are related to each other and what strength (strong, moderate, weak, or none) this relationship is. A Spearman rank correlation measures the strength of a monotonic relationship between variables based on ranked values, this test doesn’t require normality so it stays accurate even with outliers. 

### b. 

```{r}
#| label: salinity and biomass plot
#ggplot base 
ggplot(data = salinity, #starting data frame
       aes(x = salinity_mS_cm, #makes x-axis salinity
           y = pickleweed)) + #makes y-axis biomass
  geom_point() + #adds data points
  labs(x = "Salinity (mS/cm)", #adds label to x-axis
       y = "Pickleweed biomass (g)", #adds label to y-axis
       title = "As salinity increases, so does pickleweed biomass") #adds graph title 
```


### c. 

### d. 

### e. 

### f. 

## Problem 2. Personal data 
### a. 

### b. 

## Problem 3. Affective visualization 
### a. 

### b. 

### d. 

### e. 

## Problem 4. Statistical critique 
#### a. 

### b. 

### c. 

### d. 


